---
title: "CUDSS_R_Data_Vis"
author: "ls760 and Anika"
date: "12/01/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Reminder/Disclaimer

To follow the workshop you should have already a basic knowledge of R/coding. 
Fully charged laptop or charger
Have R and Rstudio installed on your laptop




```{R load libraries and import the data}

library(tidyverse) # Load the Tidyverse package. More info at https://www.tidyverse.org/

body_fat_df <- read_csv("bodyfat.csv", col_names = TRUE) # import the Body Fat Data Frame. Source: http://www-huber.embl.de/users/klaus/BasicR/bodyfat.rda

fitler

```


```{r melt}

library(reshape2)


body_fat_melt <- body_fat_df %>% 
                 melt(id.vars=c('density', 'percent.fat','age','weight','height')) # melt is using the columns you are listing to create a uniq ID which is the column name/ID, then it stacks the set of columns in a single column (more info: https://i1.wp.com/www.studytrails.com/wp-content/uploads/2016/09/reshape.png).

```

To look how you data have chanegd you can do: 
head(body_fat_df) 
dim(body_fat_df)
and compare it to:
head(body_fat_melt)
dim(body_fat_melt)

you'll see know that the variables that are the columns in the body_fat_df are observations in the dataframe body_fat_melt in the column variable. The values for the ......


```{r facet}

plot_facet <- ggplot(body_fat_melt, mapping = aes()) + # ggplot function is creating the empty graphic 
          geom_point() +
          facet_grid()



```









